---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo=TRUE, message=F, warning=F}

#load libraries
library(ggplot2);library(plyr);

#options
options(scipen=1, digits=2)

#load data
data = read.csv("activity.csv")

#remove NA steps
dataWithoutNA = subset(data, !is.na(steps))

```
## What is mean total number of steps taken per day?
First let's prepare our data:
```{r, echo=TRUE, message=F, warning=F}
stepsPerDay = tapply(dataWithoutNA$steps , dataWithoutNA$date, FUN=sum)
stepsPerDay = stepsPerDay[!is.na(stepsPerDay)]
stepsPerDay = as.data.frame(stepsPerDay)
stepsPerDay = cbind(stepsPerDay, rownames(stepsPerDay))
colnames(stepsPerDay) = c('steps', 'date')
```

Now, let's plot a histogram.
```{r, echo=TRUE, message=F, warning=F}
binwidth = (range(stepsPerDay$steps)[2] - range(stepsPerDay$steps)[1])/10
ggplot(data=stepsPerDay, aes(steps)) + geom_histogram(binwidth=binwidth) + geom_vline(aes(xintercept=mean(steps)), color="red", linetype="dashed", size=1)
```

Let's calculate the mean and the median
```{r, echo=TRUE, message=F, warning=F}
meanStepsPerDay = mean(stepsPerDay$steps)
medianStepsPerDay = median(stepsPerDay$steps)
```

The mean is `r I(meanStepsPerDay)` and the median is `r I(medianStepsPerDay)`.

## What is the average daily activity pattern?

Let's prepare the data to plot the daily activity
```{r, echo=TRUE, message=F, warning=F}
stepsPerInterval = tapply(dataWithoutNA$steps , dataWithoutNA$interval, FUN=mean)
stepsPerInterval = as.data.frame(stepsPerInterval)
stepsPerInterval = cbind(stepsPerInterval, as.numeric(rownames(stepsPerInterval)))
colnames(stepsPerInterval) = c('steps', 'interval')
```

Let's plot the time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days
```{r, echo=TRUE, message=F, warning=F}
intervalWithMax=stepsPerInterval[which.max(stepsPerInterval$steps),2]
maxSteps = max(stepsPerInterval$steps)
ggplot(data=stepsPerInterval, aes(x=interval, y=steps)) +  geom_line() + ylab("Number of steps") + geom_vline(aes(xintercept=intervalWithMax), color="red", linetype="dashed", size=1)
```

The interval `r I(intervalWithMax)` contains the maximum number of steps, which is `r I(maxSteps)`. 

## Imputing missing values

First of all, let sum the number of missing values in the dataset
```{r, echo=TRUE, message=F, warning=F}
missingSteps = sum(is.na(data$steps))
```
We have `r I(missingSteps)` of missing values

Our strategy for filling in all of the missing values in the dataset is:
  * Find the rows with missing steps
  * If we have an average value for the interval across all days, we use this mean value
  * If there is no average for the interval, we use zero
  
Here's the code that replaces the missing values.
```{r, echo=TRUE, message=F, warning=F}
replaceNa = function(row) {
  if(is.na(row[1])) {
    meanInterval = stepsPerInterval[row[3],1]
    meanInterval = if(is.na(meanInterval)) 0 else meanInterval
    as.numeric(meanInterval)
  } else {
    as.numeric(row[1])
  }
}

dataWithReplacedNA = data
dataWithReplacedNA$steps = apply(data,1,replaceNa)
dataWithReplacedNA$oldSteps = data$steps
```

Histogram of the total number of steps taken each day
```{r, echo=TRUE, message=F, warning=F}
stepsPerDayReplacedNA = tapply(dataWithReplacedNA$steps , dataWithReplacedNA$date, FUN=sum)
stepsPerDayReplacedNA = as.data.frame(stepsPerDayReplacedNA)
stepsPerDayReplacedNA = cbind(stepsPerDayReplacedNA, rownames(stepsPerDayReplacedNA))
colnames(stepsPerDayReplacedNA) = c('steps', 'date')
binwidth = (range(stepsPerDayReplacedNA$steps)[2] - range(stepsPerDayReplacedNA$steps)[1])/20
ggplot(data=stepsPerDayReplacedNA, aes(steps)) + geom_histogram(binwidth=binwidth) + geom_vline(aes(xintercept=mean(steps)), color="red", linetype="dashed", size=1)
```

Now, let's compare the mean and the median of the dataset with replaced missing values
```{r, echo=TRUE, message=F, warning=F}
meanWithReplacedNa = mean(stepsPerDayReplacedNA$steps)
medianWithReplacedNa = median(stepsPerDayReplacedNA$steps)
```

The mean is `r I(meanWithReplacedNa)`, it used to be `r I(meanStepsPerDay)`. The new median is `r I(medianWithReplacedNa)`, it used to be `r I(medianStepsPerDay)`.
## Are there differences in activity patterns between weekdays and weekends?


```{r, echo=TRUE, message=F, warning=F}
```

```{r, echo=TRUE, message=F, warning=F}
```